---
title: "Stat400Project"
author: "JamesTondt, Zach Lees, Kyler"
date: "2023-11-29"
output: html_document
---

## Front Matter

```{r}
#Libraries
library(tidyverse)
library(readr)
library(dplyr)

#Import Data
Credit_card_fraud <- read_csv("C:/Users/kyler/Downloads/application_data.csv (1).zip")

```
# Research Question
Are females at a higher risk of defaulting on loans? Does Gender have an effect on the likelihood of credit default when controlling for their total income, credit loans, and client status for credit applications.


```{r}
# Create a new data frame and clean data
CCF2 <- 
  Credit_card_fraud %>%
  select(TARGET, CODE_GENDER, AMT_INCOME_TOTAL, AMT_CREDIT, NAME_INCOME_TYPE, NAME_EDUCATION_TYPE, NAME_FAMILY_STATUS, NAME_HOUSING_TYPE) %>%
    filter(CODE_GENDER != "XNA") %>%    # only using M and F
    na.omit(CCF2)   # remove any missing values
# create indicator variable for gender
CCF2 <- CCF2 %>%
  mutate(is_female = ifelse(CODE_GENDER == "F", 1, 0))

```

```{r}
str(CCF2)
```

```{r}
summary(CCF2)
```
# Univariate Analysis
Target is our response variable, this is whether the client had payment difficulties. For study purposes Target will determine if the client is determined to have credit fraud or not.
```{r}
# Barplot for Credit Fraud distribution
ggplot(CCF2, aes(x = factor(TARGET))) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Credit Fraud",
       x = "Defaulted or Not (1 = Yes, 0 = No)",
       y = "Count") +
  theme_minimal()

```
Based upon a quick analysis of this graph, most of the data collected shows that Target (Yes =1) holds the vast majority of responses shown for credit fraud detection. This means that most of the clients, based upon the data, collected may be subject to credit fraud and that we can continue our analysis further to see if Target will be a good response variable to our research question.


Continuing on with our research, we created a graph that shows the ditribution of Gender  (male or female) to observe the split between genders collected. 
```{r}
# Barplot for Gender distribution
ggplot(CCF2, aes(x = CODE_GENDER)) +
  geom_bar(fill = "red") +
  labs(title = "Distribution of Gender",
       x = "Gender",
       y = "Count") +
  theme_minimal()

```
Based upon the graph above we can say that about twice the amount of clients reported to be females verses males. Continuing forward in our analysis this may be something to consider before stating our results later on.


This is a total income, due to an error in the graph, there may be an incorrectly entered point. Removing any data exceeding an income of 1,000,000 would keep our data from having errors or excessive outliers. We will set a maximum income value of 1,000,000 to clean our data.
```{r}
# Histogram for Total Income
ggplot(CCF2, aes(x = AMT_INCOME_TOTAL)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black") +
  labs(title = "Distribution of Total Income",
       x = "Total Income",
       y = "Count") +
  coord_cartesian(xlim = c(0, 1000000))
# Remove income past 1,000,000
CCF2 <- CCF2 %>% filter(AMT_INCOME_TOTAL < 1000000)
```

```{r}
# Histogram for Credit Amount
ggplot(CCF2, aes(x = AMT_CREDIT)) +
  geom_histogram(binwidth = 50000, fill = "orange", color = "black") +
  labs(title = "Distribution of Credit Amount",
       x = "Credit Amount",
       y = "Count") +
  theme_minimal()

```
Comparing the graphs for Total Income and Credit amount, the graphs seem to be very similar. Thus it is fare to consider the fact that the credit amount of a loan is dependent on the house income and based upon the data is closely related to the same amount.

# Multivariate Analysis
For a categorical and quantitative variable, we want to create side-by-side boxplots and look at summary statistics.

```{r}

# Scatterplot for Credit Fraud by Total Income
ggplot(CCF2, aes(x = factor(TARGET), y = AMT_INCOME_TOTAL, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Credit Fraud by Total Income",
       y = "Total Income",
       x = "Defaulted (1 = Yes, 0 = No)",
       color = "Gender")
```
Regardless of defaulting or not, males have higher income in each respective quartile. Income doesn't appear to be a very strong predictor for determining if a client defaulted or not.

```{r}
CCF2 %>% 
  group_by(TARGET) %>%
  summarize(meanMath = mean(AMT_INCOME_TOTAL),
            medianMath = median(AMT_INCOME_TOTAL),
            sdMath = sd(AMT_INCOME_TOTAL))
```


```{r}
# Scatterplot for Credit Fraud by Credit Amount
ggplot(CCF2, aes(x = TARGET, y = AMT_CREDIT, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Credit Fraud by Credit Amount",
       y = "Credit Amount",
       x = "Defaulted (1 = Yes, 0 = No)",
       color = "Gender") +
  theme_minimal()
```
The boxplots between male and female regarding if they defaulted or not don't appear to be very different for credit amount. Credit amount may not be a strong predictor for determining if a client defaulted or not.

```{r}
CCF2 %>% 
  group_by(TARGET) %>%
  summarize(meanMath = mean(AMT_CREDIT),
            medianMath = median(AMT_CREDIT),
            sdMath = sd(AMT_CREDIT))
```


Next we create a graph to show the realtion between credit fraud and income type.
```{r}
# Barplot for Credit Fraud by Client Status
ggplot(CCF2, aes(x = NAME_INCOME_TYPE, fill = factor(TARGET))) +
  geom_bar(position = "dodge") +
  labs(title = "Credit Fraud by Client Status for Credit Applications",
       x = "Client Status",
       fill = "Credit Fraud (1 = Yes, 0 = No)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
Based upon the graph we can consider that there is a relationship between income type and credit fraud. 

For further analysis we will create a a model and test the variables' correlations to Target and form a conclusion to our research question: Are females at a higher risk of credit fraud?/Does Gender have an effect on the likelihood of credit fraud when controlling for their total income, credit loans, and client status for credit applications.