---
title: "Stat400Project"
author: "JamesTondt, Zach Lees, Kyler"
date: "2023-11-29"
output: html_document
---

## Front Matter

```{r}
#Libraries
library(tidyverse)
library(readr)
library(dplyr)
library(lme4)

#Import Data
#C:/Users/20lee/Downloads/application_data.csv/application_data.csv
Credit_card_fraud <- read_csv("C:/Users/kyler/Downloads/application_data.csv (1).zip")

```
# Part 1
# Research Data Information

Use data from:

https://www.kaggle.com/datasets/mishra5001/credit-card/data?select=application_data.csv 

https://www.kaggle.com/datasets/mishra5001/credit-card/data?select=columns_description.csv 

The data presented provides information about a client with potential credit fraud. The file
includes 122 variables that are a mix of binomial and categorical. The data provided is found on
kaggle by Mishra5001 and was updated 4 years ago.

Response: TARGET(binary)

Predictors: CODE_GENDER(binary), AMT_INCOME_TOTAL(quantitative), AMT_CREDIT(quantitative), NAME_CLIENT_TYPE(binary)

Descriptions:
TARGET: client with reported payment difficulties (this variable will represent credit fraud if
positive)

CODE_GENDER: Gender of the client

AMT_INCOME_TOTAL: Income of the client

AMT_CREDIT: Credit amount of the loan

NAME_CLIENT_TYPE: Was the client old or new when applying for the previous application

Research Question: Are females at a higher risk of defaulting?/Does Gender have an effect on
the likelihood of defaulting



# Research Question Data Analysis
Are females at a higher risk of defaulting on loans? Does Gender have an effect on the likelihood of defaulting when controlling for their total income, credit loans, and client status for credit applications.


```{r}
# Create a new data frame and clean data
CCF2 <- 
  Credit_card_fraud %>%
  select(TARGET, CODE_GENDER, AMT_INCOME_TOTAL, AMT_CREDIT, NAME_INCOME_TYPE, NAME_EDUCATION_TYPE, NAME_FAMILY_STATUS, NAME_HOUSING_TYPE) %>%
    filter(CODE_GENDER != "XNA") %>%    # only using M and F
    na.omit(CCF2)   # remove any missing values

# create indicator variable for GENDER and TARGET
CCF2 <- CCF2 %>%
  mutate(is_female = ifelse(CODE_GENDER == "F", 1, 0), TargetYes = ifelse(TARGET == 1,"Yes","No"))

```

```{r}
str(CCF2)
```

```{r}
summary(CCF2)
```
# Univariate Analysis and Multivairate Analysis

Target is our response variable, this is whether the client was reported to have payment difficulties. For study purposes TargetYes will determine if the client is determined to have defaulted (yes) or not (no).

Based upon a quick analysis of this graph, most of the data collected shows that TargetYes does not hold the vast majority of responses shown for default detection. This means that most of the clients, based upon the data, collected may not be subject to defaulting and that we can continue our analysis further to see if Target will be a good response variable to our research question.

```{r}
# Barplot for Credit Fraud distribution
ggplot(CCF2, aes(x = TargetYes)) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Defaults",
       x = "Reported Defaults",
       y = "Count") +
  theme_minimal()

```



Continuing on with our research, we created a graph that shows the distribution of Gender  (male or female) to observe the split between genders collected and reported defaults.

Based upon the graph, we can say that about twice the amount of clients reported to be females verses males had no reported default. However, from further observation, we can see that females have a slightly higher count than males for reported default.s. Continuing forward in our data analysis we will observe other relationships between the variables.

```{r}
# Barplot for Gender distribution
ggplot(CCF2, aes(x = CODE_GENDER, fill = CODE_GENDER)) +
  geom_bar() +
  facet_wrap(vars(TargetYes))+
  labs(title = "Distribution of Gender",
       x = "Gender",
       y = "Count",
       color = "Gender") +
  theme_minimal()

```



Next we will oserve the relationship between total income and gender. Due to an error in the graph, there may be an incorrectly entered point. Removing any data exceeding an income of 1,000,000 would keep our data from having errors or excessive outliers. We will set a maximum income value of 1,000,000 to clean our data.

Looking at the data comparing total income and the amount of credit fraud reported by gender, we can see that females typically show a pattern to have more reported defaults than the males do with the same to similar total incomes. Looking at this graph we have to keep in mind that more females than males were also surveyed, so this graph alone is not enough to provide enough evidence to support our research statement.

```{r}
# Histogram for Total Income
ggplot(CCF2, aes(x = AMT_INCOME_TOTAL, fill = CODE_GENDER)) +
  geom_histogram(binwidth = 40000) +
  facet_wrap(vars(TargetYes))+
  labs(title = "Distribution of Total Income",
       x = "Total Income",
       y = "Count",
       color = "Gender") +
  coord_cartesian(xlim = c(0, 1000000))
# Remove income past 1,000,000
CCF2 <- CCF2 %>% filter(AMT_INCOME_TOTAL < 1000000)
```


Comparing the graphs for Total Income and Credit amount, the graphs seem to be very similar. Thus it is fare to consider the fact that the credit amount of a loan is dependent on the house income and based upon the data is closely related to the same amount.

```{r}
# Histogram for Credit Amount
ggplot(CCF2, aes(x = AMT_CREDIT)) +
  geom_histogram(binwidth = 100000, fill = "orange", color = "black") +
  labs(title = "Distribution of Credit Amount",
       x = "Credit Amount",
       y = "Count") +
  theme_minimal()

```



For a categorical and quantitative variable, we want to create side-by-side boxplots and look at summary statistics.

Interestingly we have found that males have higher income on average in each respective quarterly. Income doesn't appear to be a very strong predictor for determining if a client defaulted or not.

```{r}

# Scatterplot for Credit Fraud by Total Income
ggplot(CCF2, aes(x = TargetYes, y = AMT_INCOME_TOTAL, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Defaults by Total Income",
       y = "Total Income",
       x = "Default Depiction",
       color = "Gender")
```


```{r}
CCF2 %>% 
  group_by(TargetYes) %>%
  summarize(meanMath = mean(AMT_INCOME_TOTAL),
            medianMath = median(AMT_INCOME_TOTAL),
            sdMath = sd(AMT_INCOME_TOTAL), count = n())
```


The boxplots between male and female regarding if they have reported default or not in the data don't appear to be very different for credit amount. Credit amount may not be a strong predictor for determining if a client defaulted or not. This could be because the credit amount given is not sex dependent, so males and females recieve credit on the same rate based upon thier income.

```{r}
# Scatterplot for Credit Fraud by Credit Amount
ggplot(CCF2, aes(x = TargetYes, y = AMT_CREDIT, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Defaults by Credit Amount",
       y = "Credit Amount",
       x = "Default Depiction",
       color = "Gender") +
  theme_minimal()
```


```{r}
CCF2 %>% 
  group_by(TargetYes) %>%
  summarize(meanMath = mean(AMT_CREDIT),
            medianMath = median(AMT_CREDIT),
            sdMath = sd(AMT_CREDIT))
```


Next we create a graph to show the realtion between credit fraud and income type.

Based upon the graph we can consider that there is a relationship between income type and defaulting. 

```{r}
# Barplot for Credit Fraud by Client Status
ggplot(CCF2, aes(x = NAME_INCOME_TYPE, fill = factor(TargetYes))) +
  geom_bar(position = "dodge") +
  labs(title = "Default by Client Status for Credit Applications",
       x = "Client Status",
       y = "Count",
       fill = "Defaults") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


For further analysis we will create a a model and test the variables' correlations to TargetYes and form a conclusion to our research question: Are females at a higher risk of credit fraud?/Does Gender have an effect on the likelihood of credit fraud when controlling for their total income, credit loans, and client status for credit applications.

# Analysis


We want to analyze the model including all of our variables versus the reduced model, which doesn't include gender. 

Since the difference between these two is only one variable, we can either do a Wald Test or a Drop-in-Deviance Test. We will do a Drop-in-Deviance Test.

```{r}
#makes models
model <- glm(TARGET ~ AMT_CREDIT + AMT_INCOME_TOTAL + NAME_INCOME_TYPE + CODE_GENDER, family = binomial, data = CCF2)
model_reduced <- glm(TARGET ~ AMT_CREDIT + AMT_INCOME_TOTAL + NAME_INCOME_TYPE, family = binomial, data = CCF2)

#compares models
anova(model_reduced, model, test = "Chisq")
summary(model)
summary(model_reduced)
```

1. Hypotheses
- Null Hypothesis: $H_0:  \beta_{GENDER} = 0$
Implied model: $$log(\frac{p_i}{1-p_i}) = \beta_{0} + \beta_{1} x_{i,CREDIT} + \beta_{2} x_{i,INCOME_TOTAL} + \beta_{3} x_{i,INCOME_TYPE} + \epsilon_{i}$$
- Alternative Hypothesis: $H_1: \beta_{GENDER} \ne 0$ 
Implied model: $$log(\frac{p_i}{1-p_i}) = \beta_{0} + \beta_{1} x_{i,CREDIT} + \beta_{2} x_{i,INCOME_TOTAL} + \beta_{3} x_{i,INCOME_TYPE} + \beta_{4} x_{i,GENDER} + \epsilon_{i}$$

2. Test Statistic: Drop-in-deviance = 756.44

3. p-value : $p =< 2.2e-16$

4. Decision: 
Reject $H_0$. Since the p-value < 0.05, reject $H_0$.

5. Conclusion: 
We will use the model with the variable for gender.

Looking at our model, we can see that the gender being male is positive. This means that the gender being female is a negative predictor. A negative predictor lowers our Target value, which represents the likelihood of a client defaulting or not. If someone is female, they have a lower likelihood of defaulting on a payment.

We are 95% confidence that for CODE_GENDERM (Male), the defaults will increase by between 1.431200 and 1.511436, assuming all other variables are controlled in this model. Males have a greater odds of defaulting compared to females.

```{r}
#CI
exp(confint(model, "CODE_GENDERM", level = 0.95))
```

From backward selection, we can see that none of the predictors we have are eliminated. This means that all of them are statistically significant. This reaffirms our drop-in-deviance test, where we were testing if a client being male or female was significant for determining defaults.

```{r}
#Perform backward selection
stats::step(model, 
            scope = list(lower=~1, upper=model), 
            data = CCF2, 
            direction = 'backward')
```



# Part 2

Research Question: Does the association between meaning and reaction time differ based on old age.

Response: Reaction time (rt_sec)

Level 1 observational unit: Reaction
Level 2 observational unit: Participant

Level 1 covariates: meaning, salience
Level 2 covariates: sex, age, oldage, yrs65

Fixed Effects: meaning, oldage
Random Effects: id


```{r}
reaction <- read.csv("https://raw.githubusercontent.com/mkshaw/learn-mlms/master/data/hoffman2007.csv")
```


# Univariate Analysis

Many of the reaction times are very quick, withing the first 0 to 5 seconds, and it is right skewed. 

```{r}
# analysis for rt_sec
ggplot(reaction, aes(x = rt_sec)) +
  geom_histogram(binwidth = 0.1) +
  labs(title = "Distribution of Reaction Time",
       x = "Reaction Time (rt_sec)",
       y = "Count")
```

This is a numerical representation of how useful the picture is. It is interesting to note that it is not an equal spread, it has large amounts in 0 and 4.

```{r}
#analysis for meaning
ggplot(reaction, aes(x = meaning)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Meaning",
       x = "Meaning",
       y = "Count")
```

There are more drivers we have under the age of 40 than over the age of 40.

```{r}
#analysis for oldage
ggplot(reaction, aes(x = factor(oldage, levels = c(0, 1), labels = c("Less than 40", "Over 40")))) +
  geom_bar(fill = "pink", color = "black") +
  labs(title = "Distribution of Old Age",
       x = "Old or Not",
       y = "Count")
```

There are more females than males in this dataset.

```{r}
#analysis for sex
ggplot(reaction, aes(x = factor(sex, levels = c(0, 1), labels = c("Male", "Female")))) +
  geom_bar() +
  labs(title = "Distribution of Sex",
       x = "Sex",
       y = "Count")
```

The age of participants is bimodial. There are many who are very young, newer drivers in their teens and twenties and large amounts of drivers within the 65 to 80 age range. This was likely done to compare the two age groups and to get large differences by using such a drastic difference in ages.

```{r}
#analysis for age
ggplot(reaction, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "lightgrey", color = "black") +
  labs(title = "Distribution of Age",
       x = "Age",
       y = "Count")
```


## Bivariate Analysis

When comparing the groups of young and old, we see that each quartile of old has longer reaction times compared to the young boxplot. There is also more variance between the old group compared to the young group.

```{r}
#old vs young reaction time
ggplot(reaction, aes(x = factor(oldage, labels = c("Young", "Old")), y = rt_sec, fill = factor(oldage))) +
  geom_boxplot() +
  labs(title = "Reaction Time vs. Old Age",
       x = "Age Group",
       y = "Reaction Time (rt_sec)",
       fill = "Age Group")
```

Comparing sex with reaction time doesn't appear to be very useful, the impact of sex doesn't seem too impactful. Female does seem to have slightly more variance between it's quartiles, which may suggest they have slightly lower reaction speeds on average; however, the two groups, male and female appear very similar.

```{r}
#sex reaction time
ggplot(reaction, aes(x = factor(sex, levels = c(0, 1), labels = c("Male", "Female")), y = rt_sec, fill = factor(sex))) +
  geom_boxplot() +
  labs(title = "Reaction Time vs. Sex",
       x = "Sex",
       y = "Reaction Time (rt_sec)",
       fill = "Sex")
```

Salience represents how attention grabbing the picture is. As we would expect, the least attention grabbing pictures have some of the highest reaction times, while most of the more attention grabbing pictures have lower reaction times, in general.

```{r}
#salience reaction time
ggplot(reaction, aes(x = factor(salience), y = rt_sec, fill = factor(salience))) +
  geom_boxplot() +
  labs(title = "Reaction Time vs. Salience",
       x = "Salience",
       y = "Reaction Time (rt_sec)",
       fill = "Salience")
```
## Analysis

```{r}
#his example code pulled form lecture 21

model <- lmer(rt_sec ~ meaning + oldage + meaning:oldage + (1 + oldage | id),
               REML = FALSE,
               data = reaction)
summary(model)
anova(model)
```

## Example 4
```{r}
model.reduced <- lmer(rt_sec ~ meaning + oldage + meaning:oldage + (1 | id),
               REML = FALSE,
               data = reaction)
summary(model.reduced)
coef(model.reduced)
anova(model,model.reduced)
```

