---
title: "Stat400Project"
author: "JamesTondt, Zach Lees, Kyler"
date: "2023-11-29"
output: html_document
---

## Front Matter

```{r}
#Libraries
library(tidyverse)
library(readr)
library(dplyr)

#Import Data
Credit_card_fraud <- read_csv("C:/Users/20lee/Downloads/application_data.csv/application_data.csv")

```
# Research Question
Are females at a higher risk of defaulting on loans? Does Gender have an effect on the likelihood of credit default when controlling for their total income, credit loans, and client status for credit applications.


```{r}
# Create a new data frame and clean data
CCF2 <- 
  Credit_card_fraud %>%
  select(TARGET, CODE_GENDER, AMT_INCOME_TOTAL, AMT_CREDIT, NAME_INCOME_TYPE, NAME_EDUCATION_TYPE, NAME_FAMILY_STATUS, NAME_HOUSING_TYPE) %>%
    filter(CODE_GENDER != "XNA") %>%    # only using M and F
    na.omit(CCF2)   # remove any missing values
# create indicator variable for gender
CCF2 <- CCF2 %>%
  mutate(is_female = ifelse(CODE_GENDER == "F", 1, 0))

```

```{r}
str(CCF2)
```

```{r}
summary(CCF2)
```
# Univariate Analysis
Target is our response, this is whether the client had payment difficulties.
```{r}
# Barplot for Credit Fraud distribution
ggplot(CCF2, aes(x = factor(TARGET))) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Credit Fraud",
       x = "Defaulted or Not (1 = Yes, 0 = No)",
       y = "Count") +
  theme_minimal()

```

```{r}
# Barplot for Gender distribution
ggplot(CCF2, aes(x = CODE_GENDER)) +
  geom_bar(fill = "red") +
  labs(title = "Distribution of Gender",
       x = "Gender",
       y = "Count") +
  theme_minimal()

```
This is a total income, due to an error in the graph, there may be an incorrectly entered point. Removing any data exceeding an income of 1,000,000 would keep our data from having errors or excessive outliers. We will set a maximum income value of 1,000,000 to clean our data.
```{r}
# Histogram for Total Income
ggplot(CCF2, aes(x = AMT_INCOME_TOTAL)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black") +
  labs(title = "Distribution of Total Income",
       x = "Total Income",
       y = "Count") +
  coord_cartesian(xlim = c(0, 1000000))
# Remove income past 1,000,000
CCF2 <- CCF2 %>% filter(AMT_INCOME_TOTAL < 1000000)
```

```{r}
# Histogram for Credit Amount
ggplot(CCF2, aes(x = AMT_CREDIT)) +
  geom_histogram(binwidth = 50000, fill = "orange", color = "black") +
  labs(title = "Distribution of Credit Amount",
       x = "Credit Amount",
       y = "Count") +
  theme_minimal()

```
# Multivariate Analysis
For a categorical and quantitative variable, we want to create side-by-side boxplots and look at summary statistics.

```{r}

# Scatterplot for Credit Fraud by Total Income
ggplot(CCF2, aes(x = factor(TARGET), y = AMT_INCOME_TOTAL, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Credit Fraud by Total Income",
       y = "Total Income",
       x = "Defaulted (1 = Yes, 0 = No)",
       color = "Gender")
summary()
```


```{r}
CCF2 %>% 
  group_by(TARGET) %>%
  summarize(meanMath = mean(AMT_INCOME_TOTAL),
            medianMath = median(AMT_INCOME_TOTAL),
            sdMath = sd(AMT_INCOME_TOTAL))
```


```{r}
# is this meant to be different from the one above?
# Scatterplot for Credit Fraud by Credit Amount
ggplot(CCF2, aes(x = TARGET, y = AMT_CREDIT, color = CODE_GENDER)) +
  geom_boxplot() +
  labs(title = "Credit Fraud by Credit Amount",
       y = "Credit Amount",
       x = "Defaulted (1 = Yes, 0 = No)",
       color = "Gender") +
  theme_minimal()
```

```{r}
CCF2 %>% 
  group_by(TARGET) %>%
  summarize(meanMath = mean(AMT_CREDIT),
            medianMath = median(AMT_CREDIT),
            sdMath = sd(AMT_CREDIT))
```



```{r}
# Barplot for Credit Fraud by Client Status
ggplot(CCF2, aes(x = NAME_INCOME_TYPE, fill = factor(TARGET))) +
  geom_bar(position = "dodge") +
  labs(title = "Credit Fraud by Client Status for Credit Applications",
       x = "Client Status",
       fill = "Credit Fraud (1 = Yes, 0 = No)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

